---
title: "Analyzing sentiments"
author: "Tero Jalkanen"
date: "2023-01-14"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

## Packages -----------

library(readxl)
library(here)
library(tidyverse)
library(janitor)
library(skimr)

## Set file paths
here::i_am("text_analysis.Rmd")

```

## Data 

* The data contains 266 sentences which have been assessed as being `positive`, `negative` or `neutral`.

* We want to explore the data and see if it is useful for sentiment analysis.

## General properties of the data

* There are no missing entries

```{r}

text_df <- readxl::read_excel(path = here("data", "sentences_with_sentiment.xlsx")) %>% 
  # move sentiments under same column
  mutate(sentiment = case_when(
    Positive == 1 ~ "Positive",
    Negative == 1 ~ "Negative",
    Neutral == 1 ~ "Neutral",
    # in case of missing sentiment
    TRUE ~ "missing"
  )) %>% 
  janitor::clean_names()

# Let's take a look at the data
skim(text_df)

```

## The balance of sentiments


```{r}

text_df %>% 
  group_by(sentiment) %>% 
  summarise(n = n()) %>% 
  mutate(prercentage = 100*n/length(text_df$sentiment)) %>% 
  knitr::kable(digits = 2, caption = "The sentiment labels for the example sentences")

```

## Unique sentences

Check if there are duplicate sentences.

```{r}
## Find duplicate sentences

doubles <- duplicated(text_df$sentence)

#double check this
text_df[doubles,] %>% 
  group_by(sentence, sentiment) %>% 
  summarise(n = n())

```


## Slide with Bullets

- Bullet 1
- Bullet 2
- Bullet 3

## Slide with R Output

```{r cars, echo = TRUE}
summary(cars)
```

## Slide with Plot

```{r pressure}
plot(pressure)
```

